---
output:
  pdf_document: default
  html_document: default
---

# import & explore data

```r
library(foreign)
library(haven)
library(rmarkdown)
library(ggplot2)

physfunc <- read_dta("~/Desktop/physfunc.dta")
View(physfunc)

str(physfunc)
colnames(physfunc)
    # "id"   "occ"  "female" "grade"  "age50"  "phf"  
unique(physfunc$occ) # 6 measurement periods
sum(is.na(physfunc)) # there are 18189 NA's in this dataset
summary(physfunc) 
    # females make-up about 31% of participants 
    # mean age is 6.647 & median age is 6.621
    # mean phf is 50.284 & median is 52. 620
```

# create a new data frame with just women in it
```r
library(magrittr)
library(tidyverse)
library(reshape2)

Women <- physfunc %>% 
    filter(female == 1) %>% 
    na.omit()

View(Women) # only 12451 objects in "Women" data frame
```

# no longer need the gender column - remove it and NAs
```r
Women_2 <- Women %>%
    select(-female) %>% 
    na.omit()
View(Women_2)
```


###Questions####
# 1) Obtain summary statistics for phf and age50 at each occasion. 
```r

install.packages("lavaan")    
library(lavaan)
library(reshape2)

multi.fun <- function(x) {
    c(mean = mean(x), median = median(x), sd = sd(x), var = var(x), min = min(x), 
      max = max(x), range = range(x), quantile = quantile(x))
}

occ_1 <- Women_2 %>% 
    filter(occ == "1") %>% 
    select(-id, -occ, - grade) %>% 
    sapply(multi.fun)
occ_1

occ_2 <- Women_2 %>% 
    filter(occ == "2") %>% 
    select(-id, -occ, - grade) %>% 
    sapply(multi.fun)
occ_2

occ_3 <- Women_2 %>% 
    filter(occ == "3") %>% 
    select(-id, -occ, - grade) %>% 
    sapply(multi.fun)
occ_3

occ_4 <- Women_2 %>% 
    filter(occ == "4") %>% 
    select(-id, -occ, - grade) %>% 
    sapply(multi.fun)
occ_4

occ_5 <- Women_2 %>% 
    filter(occ == "5") %>% 
    select(-id, -occ, - grade) %>% 
    sapply(multi.fun)
occ_5

occ_6 <- Women_2 %>% 
    filter(occ == "6") %>% 
    select(-id, -occ, - grade) %>% 
    sapply(multi.fun)
occ_6
```

    # mean age increases and mean phf decreases over time

# as expected age increases by occ, and phf decreases slightly
# try to grade as a thrid dimension 

```r
library(ggplot2)
```

```{r echo = FALSE}
ggplot(Women_2, aes(x = occ, y = phf, group = occ, fill = occ)) + 
    geom_boxplot() # fit a line?
```

```{r echo = FALSE}
ggplot(Women_2, aes(x = occ, y = age50, group = occ, fill = occ)) + 
    geom_boxplot() # fit a line?
```
# ggplot(Women_2, aes(x = age50, y = phf, color = "grade")) + geom_point()



# 2) Plot physical functioning trajectories by age for the first 5 individuals
# in the dataset

# the first five id numbers are not 1:5, so find out what they are
```r
head(unique(Women_2$id)) 
    # the first six id numbers are: 1  2  3  7  9 14


first_5 <- Women_2[Women_2$id < 14, ]
View(first_5)
```
```{r echo = FALSE}
ggplot(first_5, aes(x=age50, y=phf, group = id, colour = id)) +
    geom_line()
    # generally, as age increases physical function is very decreasing

plot(first_5)  
```

# 3) Fit a simple random effects model for phf with no explanatory variables 
# (i.e. only the constant term). What is the intra-individual correlation? 
```r
install.packages("lme4")
install.packages("arm")
library(lme4)
library(arm)
```
    # think of individual as the random variable because phf and age varies 
    # differently from person-to-person over time

# fit model with a random intercept (individual represented by id number in the dataset)
```r
model_1 <- lmer(phf ~ (1 | id), data = Women_2, REML = FALSE)
summary(model_1)
```

        # Linear mixed model fit by maximum likelihood  ['lmerMod']
        # Formula: phf ~ (1 | id)
        # Data: Women_2
        # 
        # AIC      BIC   logLik deviance df.resid 
        # 87077.6  87099.9 -43535.8  87071.6    12448 
        # 
        # Scaled residuals: 
        #     Min      1Q  Median      3Q     Max 
        # -4.6796 -0.4190  0.1098  0.5159  3.9693 
        # 
        # Random effects:
        #     Groups   Name        Variance Std.Dev.
        # id       (Intercept) 61.2     7.823   
        # Residual             41.5     6.442   
        # Number of obs: 12451, groups:  id, 2697
        # 
        # Fixed effects:
        #     Estimate Std. Error t value
        # (Intercept)  47.8880     0.1637   292.6


# 4) Fit a random intercepts growth model which allows an individual’s physical 
# functioning to change linearly with age. 

#Fitting the  model with the random intercept
```r
model_2 <- lmer(phf ~ age50+(1 | id), data = Women_2, REML = FALSE)
summary(model_2)
```

        # Linear mixed model fit by maximum likelihood  ['lmerMod']
        # Formula: phf ~ age50 + (1 | id)
        # Data: Women_2
        # 
        # AIC      BIC   logLik deviance df.resid 
        # 86467.5  86497.2 -43229.8  86459.5    12447 
        # 
        # Scaled residuals: 
        #     Min      1Q  Median      3Q     Max 
        # -4.5153 -0.4393  0.1078  0.5267  3.7917 
        # 
        # Random effects:
        #     Groups   Name        Variance Std.Dev.
        # id       (Intercept) 58.81    7.669   
        # Residual             39.43    6.279   
        # Number of obs: 12451, groups:  id, 2697
        # 
        # Fixed effects:
        #     Estimate Std. Error t value
        # (Intercept)  49.6852     0.1757  282.86
        # age50        -0.2655     0.0106  -25.05
        # 
        # Correlation of Fixed Effects:
        #     (Intr)
        # age50 -0.409


# 5)  Fit a random slopes model that allows the rate of change in physical 
# functioning to vary among individuals. 

#Fitting the  model with the random intercept and random slope
```r
model_3 <- lmer(phf ~ age50+(1 | id) + (0 + age50 | id), data = Women_2, REML = FALSE)
summary(model_3)
```
        # Linear mixed model fit by maximum likelihood  ['lmerMod']
        # Formula: phf ~ age50 + (1 | id) + (0 + age50 | id)
        # Data: Women_2
        # 
        # AIC      BIC   logLik deviance df.resid 
        # 86152.7  86189.8 -43071.3  86142.7    12446 
        # 
        # Scaled residuals: 
        #     Min      1Q  Median      3Q     Max 
        # -4.3785 -0.4142  0.1063  0.5026  3.5006 
        # 
        # Random effects:
        #     Groups   Name        Variance Std.Dev.
        # id       (Intercept) 51.2462  7.1586  
        # id.1     age50        0.1395  0.3735  
        # Residual             34.8499  5.9034  
        # Number of obs: 12451, groups:  id, 2697
        # 
        # Fixed effects:
        #     Estimate Std. Error t value
        # (Intercept) 49.83138    0.16725  297.95
        # age50       -0.26557    0.01272  -20.88
        # 
        # Correlation of Fixed Effects:
        #     (Intr)
        # age50 -0.319


# 6) Carry out a likelihood ratio test of the null hypothesis that growth rate 
# is fixed across individuals.

# As I understand, to do a likelihood ratio test, I must compare two models
# I'm comparing two models - model_1 and model_2
```r
anova(model_1, model_2)
```
    # Data: Women_2
    # Models:
    #     model_1: phf ~ (1 | id)
    # model_2: phf ~ age50 + (1 | id)
    # Df   AIC   BIC logLik deviance  Chisq Chi Df Pr(>Chisq)    
    # model_1  3 87078 87100 -43536    87072                             
    # model_2  4 86468 86497 -43230    86460 612.09      1  < 2.2e-16 ***
    #     ---
    #     Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1


# 7) Produce a plot of that superimposes predicted phf trajectories from the 
# random intercepts and random slopes models and the observed values of phf 
# for the first 5 individuals. 
```r
library(lattice)

# This was model_3:
model_3 <- lmer(phf ~ age50+(1 | id) + (0 + age50 | id), data = Women_2, REML = FALSE)
summary(model_3)
```
```r
xyplot(fitted(model_3) ~ age50 | phf, groups=id, lwd=1, t="b", pch=19,
       data = Women_2)
```
# 8)  Fit a quadratic growth model with a fixed coefficient for age-squared. 
# (A model with a random coefficient for age-squared fails to converge.)
```r
install.packages("nlme")
library(nlme)

model_4 <- lme(phf ~ age50 + I (age50^2), random =~ age50 | id, data = Women_2)
summary(model_4)
```
        # Linear mixed-effects model fit by REML
        # Data: Women_2 
        # AIC      BIC    logLik
        # 86134.25 86186.26 -43060.13
        # 
        # Random effects:
        #     Formula: ~age50 | id
        # Structure: General positive-definite, Log-Cholesky parametrization
        # StdDev    Corr  
        # (Intercept) 7.0762752 (Intr)
        # age50       0.3576407 0.063 
        # Residual    5.9142383       
        # 
        # Fixed effects: phf ~ age50 + I(age50^2) 
        # Value  Std.Error   DF   t-value p-value
        # (Intercept) 49.80392 0.16548872 9752 300.95055       0
        # age50       -0.18133 0.01802704 9752 -10.05856       0
        # I(age50^2)  -0.00643 0.00099026 9752  -6.48870       0
        # Correlation: 
        #     (Intr) age50 
        # age50      -0.214       
        # I(age50^2)  0.012 -0.719
        # 
        # Standardized Within-Group Residuals:
        #     Min         Q1        Med         Q3        Max 
        # -4.3644586 -0.4121647  0.1068419  0.5024621  3.4697630 
        # 
        # Number of Observations: 12451
        # Number of Groups: 2697 



# Compare this model to the linear growth model using a likelihood ratio test. 

# in order to compare model_3 and model_4
```r
model_3.2 <- lme(phf ~ age50, random =~ age50 | id, data = Women_2)
summary(model_3.2)
```
        # Linear mixed-effects model fit by REML
        # Data: Women_2 
        # AIC      BIC    logLik
        # 86162.1 86206.68 -43075.05
        # 
        # Random effects:
        #     Formula: ~age50 | id
        # Structure: General positive-definite, Log-Cholesky parametrization
        # StdDev    Corr  
        # (Intercept) 7.0854582 (Intr)
        # age50       0.3650564 0.053 
        # Residual    5.9161572       
        # 
        # Fixed effects: phf ~ age50 
        # Value  Std.Error   DF   t-value p-value
        # (Intercept) 49.82151 0.16577436 9753 300.53813       0
        # age50       -0.26546 0.01262867 9753 -21.02015       0
        # Correlation: 
        #     (Intr)
        # age50 -0.298
        # 
        # Standardized Within-Group Residuals:
        #     Min         Q1        Med         Q3        Max 
        # -4.3825737 -0.4152879  0.1058552  0.5025557  3.5242316 
        # 
        # Number of Observations: 12451
        # Number of Groups: 2697 
```r
anova(model_4, model_3.2)
```
        # Model df      AIC      BIC    logLik   Test  L.Ratio p-value
        # model_4       1  7 86134.25 86186.26 -43060.13                        
        # model_3.2     2  6 86162.10 86206.68 -43075.05 1 vs 2 29.84889  <.0001
        # Warning message:
        #     In anova.lme(model_4, model_3.2) :
        #     fitted objects with different fixed effects. REML comparisons are not meaningful.
